Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-13 paper.tex
up to 2.1

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. In \emph{tight integration},
  the workflow engine shares software components with the science
  gateway. In \emph{service invocation}, the engine is isolated and
  invoked through a specific software interface. In \emph{task
    encapsulation}, the engine is wrapped as a computing task executed
  on the infrastructure. In the \emph{pool model}, the engine is
  bundled in an agent that connects to a central pool to fetch and
  execute workflows. In \emph{nested workflows}, the engine is
  integrated as a child process of another engine. In \emph{workflow
    import}, the engine is integrated through workflow language
  conversion. We describe and evaluate these architectures with
  metrics that measure integration complexity, robustness,
  extensibility, scalability and support for meta-workflows and
  fine-grained debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}. Results provide insights for
  science gateway architects and workflow engine developers. Tight
  integration and task encapsulation are the easiest to integrate and
  are most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \closednote{Naj}{I find this abstract to be vague, it can be made
    more concise and poignant}{Tristan}{Added a description of the
    architectures, following @pbellec's suggestion.}
\end{abstract}

@@

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years; as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of 
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), \closednote{Naj}{I think and is redundant here, you can replace is with a comma}{Tristan}{yes!} using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine, while other ones are more general and host applications
executed by different types of engines. %Some gateways may not even use any
%workflow engine and focus on the execution of simple tasks.

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in
Figure~\ref{fig:notations}. Table~\ref{table:system-classification}
summarizes the classification of a few systems by architecture.

@@

The interactions involved in the architectures are described below and
labeled as in Figure~\ref{fig:architectures}.

@@

\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable

@@

  well-structured and relatively simple workflow languages such as between GWorkflowDL
  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the

@@

>>>>

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. In \emph{tight integration},
  the workflow engine shares software components with the science
  gateway. In \emph{service invocation}, the engine is isolated and
  invoked through a specific software interface. In \emph{task
    encapsulation}, the engine is wrapped as a computing task executed
  on the infrastructure. In the \emph{pool model}, the engine is
  bundled in an agent that connects to a central pool to fetch and
  execute workflows. In \emph{nested workflows}, the engine is
  integrated as a child process of another engine. In \emph{workflow
    import}, the engine is integrated through workflow language
  conversion. We describe and evaluate these architectures with
  metrics for assessment of integration complexity, robustness,
  extensibility, scalability and support for meta-workflows and
  fine-grained debugging\closednote{Naj}{Please be precise, what are
    they?}{Tristan}{Done, thanks}. Results provide insights for
  science gateway architects and workflow engine developers. Tight
  integration and task encapsulation are the easiest to integrate and
  the most robust. Extensibility is equivalent in most
  architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. \closednote{Naj}{I find this abstract to be vague, it can be made
    more concise and poignant}{Tristan}{Added a description of the
    architectures, following @pbellec's suggestion.}
\end{abstract}

@@

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years, as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of 
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), \closednote{Naj}{I think and is redundant here, you can replace is with a comma}{Tristan}{yes!} using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} 
Some workflow engines, 
usually from the e-Science community, 
may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine, while others are more general and host applications
executed by different types of engines. %Some gateways may not even use any
%workflow engine and focus on the execution of simple tasks.

@@

Architectures to integrate workflow engines in science gateways are
presented in Figure~\ref{fig:architectures} using the graphical
notations defined in
Figure~\ref{fig:notations}. Table~\ref{table:system-classification}
summarizes the classification of a few systems by architecture.

@@

The interactions involved in the architectures are described below and
labeled (a,b,c, ...) as in Figure~\ref{fig:architectures}.

@@

\item Workflow conversion: translation from one workflow language into
  another. This interaction may not be available or possible

@@

  translation between well-structured and relatively simple workflow languages such as GWorkflowDL
  and Scufl~\cite{OLAB-09}, and the 4 languages that were involved in the

@@

 \note{Silvia}{Should we add here the 4?
 	and for translation among the 5 workflow systems in the SHIWA project: Askalon, MOTEUR, Taverna, Triana and WS-PGRADE)} 
